A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-01 main.tex
Changing references to Lasso to 'ordinary Lasso'

\begin{equation}
\label{eqn:rtY_distn}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX \bbeta, \bI),
\end{equation}
and thus a regular \anna{vanilla, naive, regular?} \pb{how about ``ordinary''} Lasso model can be fit on the transformed data, ($\bV^{-\frac{1}{2}}\bX, \bV^{-\frac{1}{2}}\by$). \anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?} \pb{I don't think we need to}

@@

%------------------------------%
\subsection{PC-Lasso and LMM-Lasso outperform Lasso in the presence of environmental confounding}
%------------------------------%

We begin by comparing the ordinary lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding and potential estimation bias in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1.1]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean true positive rate and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Semi-Empirical data sets than in the 1D Linear Admixture data. Again, this is due to the reduced proportion of environmental confounding that is in the column space of $\boldsymbol{P}_{\bX}$, which results in a lesser degree of confounding of the SNP-phenotype associations.

@@

We have detailed the assumptions made by PC-Lasso and LMM-Lasso, which model confounding influences as fixed and random effects, respectively. We have also described the ways in which simulation schemes to evaluate these methods may fail to accurately capture the effects of environmental heterogeneity. In turn, we propose and utilize a simulation approach in which the estimation bias introduced by environmental heterogeneity may be evaluated. We then compare the estimation accuracy of a \anna{naive} Lasso, PC-Lasso, and LMM-Lasso in various confounding scenarios.

Unsurprisingly, we find that PC-Lasso and LMM-Lasso outperform the \anna{naive} Lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-Lasso may hinder estimation accuracy and perform more poorly than even the \anna{naive} Lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-Lasso outperforms PC-Lasso when (1) fine subpopulation structure is present and (2) when confounding effects are nonlinear and discordant in nature. In the presence of coarse subpopulation structure which corresponds to concordant, dichotomous environmental effects, PC-Lasso may outperform LMM-Lasso in estimation accuracy. However, LMM-Lasso is more robust to subpopulation structure as well as nonlinear and discordant environmental effects. Therefore, LMM-Lasso is a more robust method for practitioners seeking to obtain unbiased estimates when the exact nature of the population structure and corresponding environmental heterogeneity are unknown.

>>>>

\begin{equation}
\label{eqn:rtY_distn}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX \bbeta, \bI),
\end{equation}
and thus an ordinary Lasso model can be fit on the transformed data, ($\bV^{-\frac{1}{2}}\bX, \bV^{-\frac{1}{2}}\by$).

@@

%------------------------------%
\subsection{PC-Lasso and LMM-Lasso outperform the ordinary Lasso in the presence of environmental confounding}
%------------------------------%

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding and potential estimation bias in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1.1]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}
\anna{talk about karl rohe stuff here - if anywhere}
In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean true positive rate and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Semi-Empirical data sets than in the 1D Linear Admixture data. Again, this is due to the reduced proportion of environmental confounding that is in the column space of $\boldsymbol{P}_{\bX}$, which results in a lesser degree of confounding of the SNP-phenotype associations.

@@

We have detailed the assumptions made by PC-Lasso and LMM-Lasso, which model confounding influences as fixed and random effects, respectively. We have also described the ways in which simulation schemes to evaluate these methods may fail to accurately capture the effects of environmental heterogeneity. In turn, we propose and utilize a simulation approach in which the estimation bias introduced by environmental heterogeneity may be evaluated. We then compare the estimation accuracy of an ordinary Lasso, PC-Lasso, and LMM-Lasso in various confounding scenarios.

Unsurprisingly, we find that PC-Lasso and LMM-Lasso outperform the ordinary Lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-Lasso may hinder estimation accuracy and perform more poorly than even the ordinary Lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-Lasso outperforms PC-Lasso when (1) fine subpopulation structure is present and (2) when confounding effects are nonlinear and discordant in nature. In the presence of coarse subpopulation structure which corresponds to concordant, dichotomous environmental effects, PC-Lasso may outperform LMM-Lasso in estimation accuracy. However, LMM-Lasso is more robust to subpopulation structure as well as nonlinear and discordant environmental effects. Therefore, LMM-Lasso is a more robust method for practitioners seeking to obtain unbiased estimates when the exact nature of the population structure and corresponding environmental heterogeneity are unknown.
